\chapter{Discussion}

F2J is a new born programming language. While we using it to build the parser and pretty printer library, we have encountered a lot of problems. Most of them were solved with the members in HKU programming group, but still could be improved:

\begin{itemize}
\item Module System

Currently F2J only supports standalone compilation, which compiles one independent file at a time. This naive method makes it impossible to reuse written codes, leading to hardship in managing the project as a whole.

\item Exceptions

F2J does not supports exceptions, so that I cannot use those Java functions that will throw exceptions, such as I/O libraries. Currently we could make use of the power of Java interpolation of F2J by writing a bridge class in Java to call those fuctions that may throw exceptions, but that just a temporary solution.

\item Type Inferences

F2J's type parameters must be applied explicitly, which makes the code looks too verbose and hard to read.

\item Macros

F2J does not support macros now, but it is useful sometimes. For example, when I was using \texttt{Thunk} in the library, I have to construct a \texttt{Thunk} explicitly, such as

\begin{lstlisting}
let singleton[A] (x : A) : PList[A] =
    Cons[A] x (\(__: Unit) -> (Nil[A]));
\end{lstlisting}

If F2J supports macros, the lambda function could be replaced by a more expressive macro.

\item File Name too Long

When F2J's compiler translating programs to Java, a lot of nested inner classes are generated. At some point, the level of nesting becomes so deep that we get:

\begin{lstlisting}
f2j -r f2jparserandprinter.sf

f2jparserandprinter using [Naive]

Compiling to Java source code ( ./F2jparserandprinter$.java )

./F2jparserandprinter$.java:29140: error: error while writing F22105:

./F2jparserandprinter$$1L43$1L306$1L387$1L660$1L731$1L834$1L897$1L969$1L1050
$1L1197$1L3266$1L4567$1Datatype22175$1L5857$1L7241$1L9394$1L10526$1L18657$1
L18933$1L19103$1L19299$1L19465$1L19771$1L19869$1L19974$1L21818$1F22084$1F22087
$1F22090$1F22093$1F22096$1F22105.class
(File name too long)

class F22105 extends f2j.Closure
^
1 error
Error: Could not find or load main class F2jparserandprinter$
\end{lstlisting} 

To sovle this problem, a Module System can offer a perfect solution. Or with more concise counting methods, the maximum length of code can also be elongated considerably. Further, we do not regard it a good idea to rampantly generate inner classes during the compilation. We believe that a much more appropriate, efficient and elegant approach must be around the corner.

\end{itemize}